TacticToe: Learning to Prove with Tactics

### 2018-03-21 tactictoe_jv.tex
Make minor edits in Section 1

\subsection{The problem}
An ITP is a development environment for the construction of formal proofs in 
which it is possible to write a proof as a sequence of 
applications of primitive inference rules. This approach is not preferred 
by developers because proving a valuable theorem requires in general more than 
thousand primitive inferences.
The common approach, however, is to use and devise high-level tools and 
automation that abstract over useful ways of producing proof pieces.
A specific, prevalent instance of this approach is the use of \emph{tactics} 
for \emph{backward} or goal-directed proof.
Here, the ITP user operates on a proof state, initially the desired conjecture, 
and applies tactics that transform the proof state until it is solved.
Each tactic performs a sound transformation of the proof state: essentially, it 
promises that if the resulting proof state can be solved, then so can the 
starting one. This gives rise to a machine learning problem: can we learn a 
mapping from a given proof state to the next tactic (or sequence of tactics) 
that will productively advance the proof towards a solution?

@@

\paragraph{Tactics in \holfour.}
A \emph{tactic} is a function
that takes a goal (or proof obligation) and returns a list of goals (subgoals
of the original goal) together
with a \textit{validation function}.
Calling the validation function on a list of \emph{theorems}, 
corresponding to
the list of subgoals, results in a theorem corresponding to the original goal.
For example, if the list of subgoals is empty, then calling the 
validation
function on the empty list yields the originally goal sequent as a theorem.
In this way, tactics implement the plumbing of goal-directed proof
construction, transforming the conjecture by reasoning in a backward manner.
Since validation functions are only important to check the final 
proof, we
omit them during the description of the proof search algorithm. We denote by
$t(g)$ the list of goals produced by the application of a tactic $t$ to a goal
$g$.

\subsection{Contributions}
This paper extends the work described in our previous work~\cite{tgckju-lpar17}.
There we proposed the idea of a tactic-based prover based on supervised
learning guidance. And we achieved a 39\% success rate on theorems of the
\holfour standard library by running \tactictoe with a 5 seconds timeout.
The contributions of this paper and their effect on our system are:

\begin{itemize}
\item Proof recording at the tactic level is made more precise in 
Section~\ref{sec:recording}. Pattern
matching constructions, opening of submodules are supported. 
%The different
%recording steps (parsing, unfolding and replaying) are regrouped into a single
%\sml function.
\item Monte Carlo tree search (MCTS) replaces A* as our
search algorithm in Section~\ref{sec:proofsearch}. The MCTS algorithm gives a 
more balanced feedback on
the success of
each proof step by
comparing subtrees of the search tree instead of leafs. The policy and
evaluation are learned
through supervised learning.
\item Proof guidance required by MCTS is given by prediction algorithms 
through supervised learning (see 
Section~\ref{s:prediction}).
These predictors are implemented for three kind of objects: tactics, 
theorems and lists of goals.
\item The orthogonalization process that eliminates redundant tactics is
described in a precise manner in Section~\ref{sec:ortho}.
\item A tactic abstraction mechanism is introduced in 
Section~\ref{sec:synthesis}. It enables us to create
more general and flexible tactics where tactic arguments are dynamically
predicted.
\item The internal ATP \metis is complemented
by asynchronous calls to \eprover in Section~\ref{sec:atp} to help \tactictoe 
during proof search.
\item Evaluation of \tactictoe with a 60 seconds timeout achieves a 66\%
success rate on the standard library. And comparisons between 
\tactictoe and \eprover on different type of problems are reported in 
Section~\ref{s:experiments}.
\item Minimization and embellishment of the discovered proof facilitates user
interaction in Section~\ref{sec:proofdisplay}.
\end{itemize}

@@

\end{tikzpicture}
\end{center}
\caption{\label{fig:components}Relation between modules of the learning and 
proving tool chains. Functions are represented by rectangles and objects by 
ellipses.}
\end{figure}

The different components depicted in Figure~\ref{fig:components} give a 
high-level overview of the techniques necessary to generate 
tactic-level proofs by learning from recorded knowledge. From a formal library 
of ITP proofs, we extract and record tactic invocations 
in Section~\ref{sec:recording}. Based on such training examples, we
  build efficient machine learning classifiers in Section~\ref{s:prediction}.
  Modification of the knowledge base (shown as the training loop in 
  Figure~\ref{fig:components})
  achieved by orthogonalization in Section~\ref{sec:ortho} and completed by 
  an abstraction technique in Section~\ref{sec:synthesis}
  improves the quality of the predictions.
  And the preselection method described in Section~\ref{sec:presel} increases 
  the speed of the prediction algorithm.
  The learned predictors relying on this knowledge guide the proof search 
  algorithm presented in Section~\ref{sec:proofsearch}.
  The search algorithm explores different proof paths in the proof tree which 
  is defined in Section~\ref{sec:prelim}. 
  The result, if
  successful, is a verified human-level proof composed of \holfour
  tactics minimized in Section~\ref{sec:proofdisplay}. And in 
  Section~\ref{s:experiments}, we evaluate the system 
  on a large set of theorems originating from \holfour.
  The generated proofs are compared with human proofs in 
  Section~\ref{sec:case_study}.

The section order does not follow the implementation to give a description of 
the proof search algorithm sooner. This order is 
beneficial for the reader as it shows the essential of our work before 
technical details on the prediction and recording algorithms.

>>>>

\subsection{The problem}
An ITP is a development environment for the construction of formal proofs in
which it is possible to write a proof as a sequence of
applications of primitive inference rules. This approach is not preferred
by developers because proving a valuable theorem requires often many
thousands of primitive inferences.
The preferred approach is to use and devise high-level tools and
automation that abstract over useful ways of producing proof pieces.
A specific, prevalent instance of this approach is the use of \emph{tactics}
for \emph{backward} or goal-directed proof.
Here, the ITP user operates on a proof state, initially the desired conjecture,
and applies tactics that transform the proof state until it is solved.
Each tactic performs a sound transformation of the proof state: essentially, it
promises that if the resulting proof state can be solved, then so can the
initial one. This gives rise to a machine learning problem: can we learn a
mapping from a given proof state to the next tactic (or sequence of tactics)
that will productively advance the proof towards a solution?

@@

\paragraph{Tactics in \holfour.}
A \emph{tactic} is a function
that takes a goal (or proof obligation) and returns a list of goals (subgoals
of the original goal) together
with a \textit{validation function}.
Calling the validation function on a list of \emph{theorems},
corresponding to
the list of subgoals, results in a theorem corresponding to the original goal.
For example, if the list of subgoals is empty, then calling the
validation
function on the empty list yields the original goal sequent as a theorem.
In this way, tactics implement the plumbing of goal-directed proof
construction, transforming the conjecture by reasoning in a backward manner.
Since validation functions are only important to check the final
proof, we
omit them during the description of the proof search algorithm. We denote by
$t(g)$ the list of goals produced by the application of a tactic $t$ to a goal
$g$.

\subsection{Contributions}
This paper extends work described previously~\cite{tgckju-lpar17}, in which
we proposed the idea of a tactic-based prover based on supervised
learning guidance. We achieved a 39\% success rate on theorems of the
\holfour standard library by running \tactictoe with a 5 second timeout.
The contributions of this paper and their effect on our system are:

\begin{itemize}
\item Proof recording at the tactic level is made more precise
  (Section~\ref{sec:recording}). We now support pattern
matching constructions and opening SML modules.
%The different
%recording steps (parsing, unfolding and replaying) are regrouped into a single
%\sml function.
\item Monte Carlo tree search (MCTS) replaces A* as our
  search algorithm (Section~\ref{sec:proofsearch}). The MCTS algorithm gives
more balanced feedback on
the success of
each proof step by
comparing subtrees of the search tree instead of leaves. The policy and
evaluation are learned
through supervised learning.
\item Proof guidance required by MCTS is given by prediction algorithms
through supervised learning (Section~\ref{s:prediction}).
These predictors are implemented for three kind of objects: tactics,
theorems and lists of goals.
\item We introduce an orthogonalization process that eliminates redundant tactics (Section~\ref{sec:ortho}).
\item We introduce a tactic abstraction mechanism
  (Section~\ref{sec:synthesis}), which enables us to create
more general and flexible tactics by dynamically
predicting tactic arguments.
\item The internal ATP \metis is complemented
  by asynchronous calls to \eprover (Section~\ref{sec:atp}) to help \tactictoe
during proof search.
\item Evaluation of \tactictoe with a 60 seconds timeout achieves a 66\%
success rate on the standard library. Comparisons between
\tactictoe and \eprover on different type of problems are reported in
Section~\ref{s:experiments}.
\item Minimization and embellishment of the discovered proof facilitates user
  interaction (Section~\ref{sec:proofdisplay}).
\end{itemize}

@@

\end{tikzpicture}
\end{center}
\caption{\label{fig:components}Relation between modules of the learning and
proving toolchains. Functions are represented by rectangles and objects by
ellipses.}
\end{figure}

The different components depicted in Figure~\ref{fig:components} give a
high-level overview of the techniques necessary to generate
tactic-level proofs by learning from recorded knowledge. From a formal library
of ITP proofs, we extract and record tactic invocations
(Section~\ref{sec:recording}). Based on such training examples, we
  build efficient machine learning classifiers (Section~\ref{s:prediction}).
  We then modify the knowledge base (shown as the training loop in
  Figure~\ref{fig:components})
  using orthogonalization (Section~\ref{sec:ortho}) and
  an abstraction technique (Section~\ref{sec:synthesis})
  to improve the quality of the predictions.
  Additionally, the preselection method described in Section~\ref{sec:presel} increases
  the speed of the prediction algorithm.
  The learned predictors relying on this knowledge guide the proof search
  algorithm presented in Section~\ref{sec:proofsearch}.
  The search algorithm explores different proof paths in the proof tree, which
  is defined in Section~\ref{sec:prelim}.
  The result, if
  successful, is a verified human-level proof composed of \holfour
  tactics, minimized according to Section~\ref{sec:proofdisplay}.
  In
  Section~\ref{s:experiments}, we evaluate the system
  on a large set of theorems originating from \holfour.
  The generated proofs are compared with human proofs in
  Section~\ref{sec:case_study}.

The section order does not follow the implementation, because we have prioritized giving a description of
the proof search algorithm early on. We hope this order
benefits the reader by showing the essence of our work before
the technical details of the prediction and recording algorithms.
